---
title: "Machine Learning Workflow"
author: "Shannon Stiles and Parth Khatri"
date: "Due 12/12/2020"
output: word_document
---

```{r setup}

# read in libraries
library(tidyverse)
library(dplyr)
library(stringr)
library(e1071)
library(class)

# MAKE SURE YOUR KNIT DIRECTORY IS SET TO YOUR PROJECT DIRECTORY

```


# Preprocessing
```{r read_data}

training <- readRDS("./data/training_data")

testing = readRDS("./data/testing_data")

DE_genes <- readRDS("./data/DE_genes")

```


```{r subset_data}

# define columns of differentially expressed(DE) genes to subset datasets
cols <- c(DE_genes, "donor", "label")

# subset training and testing data for only DE genes and donor and label cols
training <- subset(training, select = cols)
testing <- subset(testing, select = cols)

# Define X and Y for training and testing datasets
X_train <- training%>%select(-"donor", -"label")
  
Y_train <- as.vector(t(training%>%select("label")))

X_test <- testing%>%select(-"donor", -"label")
  
Y_test <- as.vector(t(testing%>%select("label")))

```

# K-Nearest Neighbors Classification of Covid and Control Patient Cells
```{r knn_classifier}

# run KNN classification with timer
system.time(knn_classifier <- knn(X_train, X_test, Y_train, k = round(sqrt(nrow(X_train)))))

```


```{r knn_results}

# summarize KNN classification results in a confusion matrix
conf_mx <- table(knn_classifier, Y_test)
conf_mx

# calculate accuracy of the test set
accuracy <- function(x){sum(diag(x)/(sum(rowSums(x)))) * 100}
accuracy(conf_mx)

knn_sens <- conf_mx[4] / (conf_mx[4] + conf_mx[3])
knn_sens

knn_spec <- conf_mx[1] / (conf_mx[1] + conf_mx[2])
knn_spec

```


# Radial Basis Function (RBF) SVM Classification of Covid and Control Patient Cells

# Optimize C and Gamma Parameters
- C: 0.01, 0.1, 1, 10
- Gamma: 0.000001, 0.001, 1, 10

```{r opt_svm}

# initialize range of C parameters
c_range <- c(0.01, 0.1, 1, 10)

# initialize range of Gamma parameters
gamma_range <- c(0.000001, 0.001, 1, 10)


optimization_results <- data.frame("c" = numeric(),
                                   "gamma" = numeric(),
                                   "acc" = numeric(), 
                                   "sens" = numeric(),
                                   "spec" = numeric())


# iterate for donors randomly assigned to testing dataset
for (curr_c in c_range) {
  for (curr_gamma in gamma_range) {
    
    print(str_c("C: ", curr_c, " Gamma: ", curr_gamma))
    
    # run RBF SVM classification with timer
    system.time(svm_classifier <- svm(X_train, Y_train, type = "C-classification", kernel = "radial", gamma = curr_gamma, cost = curr_c))
  
    # display information about the svm classifier
    summary(svm_classifier)
    saveRDS(svm_classifier, str_c("./data/svm_opt/svm_", curr_c, "_", curr_gamma))
    
    # predict testing data
    system.time(pred <- predict(svm_classifier, X_test))
    
    # summarize KNN classification results in a confusion matrix
    svm_conf_mx <- table(pred, Y_test)

    # calculate the accuracy of the test set
    accuracy <- function(x){sum(diag(x)/(sum(rowSums(x)))) * 100}
    svm_acc <- accuracy(svm_conf_mx)
  
    # calculate the sensitivity
    svm_sens <- svm_conf_mx[4] / (svm_conf_mx[4] + svm_conf_mx[3])
  
    # calculate the specificity
    svm_spec <- svm_conf_mx[1] / (svm_conf_mx[1] + svm_conf_mx[2])
  
    curr_parameters <- data.frame(curr_c, curr_gamma, svm_acc, svm_sens, svm_spec)
    
    print(curr_parameters)
    
    optimization_results <- rbind(optimization_results, curr_parameters)
  }
}

print(optimization_results)

path <- "./data/svm_opt"

# write.csv(optimization_results, file.path(path, "SVM_Optimization_Results.csv"),row.names = TRUE)

```

# Check RBF SVM results using the default C and gamma hyperparameter values
```{r default_rbf_svm_classifier}

# run RBF SVM classification with timer
system.time(svm_classifier <- svm(X_train, Y_train, type = "C-classification", kernel = "radial", gamma = 1 / ncol(X_train), cost = 1))

```

```{r default_rbf_svm_summary}

# display information about the svm classifier
summary(svm_classifier)

```

```{r default_rbf_svm_predictions}

# predict testing data
system.time(pred <- predict(svm_classifier, X_test))

```

```{r default_rbf_svm_results}

# summarize KNN classification results in a confusion matrix
svm_conf_mx <- table(pred, Y_test)
svm_conf_mx

# calculate the accuracy of the test set
accuracy(svm_conf_mx)

# calculate sensitivity
svm_sens <- svm_conf_mx[4] / (svm_conf_mx[4] + svm_conf_mx[3])
svm_sens

# calculate specificity
svm_spec <- svm_conf_mx[1] / (svm_conf_mx[1] + svm_conf_mx[2])
svm_spec


```


# Run RBF SVM with the best hyperparameters: C = 0.1 and Gamma = 0.001
```{r rbf_svm_classifier}

# run RBF SVM classification with timer
system.time(svm_classifier <- svm(X_train, Y_train, type = "C-classification", kernel = "radial", gamma = 0.001, cost = 0.1))

```

```{r rbf_svm_summary}

# display information about the svm classifier
summary(svm_classifier)

```

```{r rbf_svm_predictions}

# predict testing data
system.time(pred <- predict(svm_classifier, X_test))

```


```{r rbf_svm_results}

# summarize KNN classification results in a confusion matrix
svm_conf_mx <- table(pred, Y_test)
svm_conf_mx

# calculate the accuracy of the test set
accuracy(svm_conf_mx)

# calculate sensitivity
svm_sens <- svm_conf_mx[4] / (svm_conf_mx[4] + svm_conf_mx[3])
svm_sens

# calculate specificity
svm_spec <- svm_conf_mx[1] / (svm_conf_mx[1] + svm_conf_mx[2])
svm_spec


```

# Summarize Cell Classification Results in ROC curves

```{r reformat_preds}

# Reformat ML algorithm predictions from factors to numeric vectors
knn_preds <- as.numeric(as.vector(knn_classifier))

svm_preds <- as.numeric(as.vector(pred))

```


```{r summary_curves}

library(ROCit)

# KNN ROC curve
knn_ROCit <- rocit(score = knn_preds, class = Y_test)
plot(knn_ROCit)

# print KNN AUC value
knn_ROCit$AUC

# SVM ROC Curve
svm_ROCit <- rocit(score = svm_preds, class = Y_test)
plot(svm_ROCit)

# print SVM AUC value
svm_ROCit$AUC

```


# Classify Covid and Control Donors based on Cell Classification

```{r create_pred_df}
# convert prediction vectors to dataframes

knn_pred_df <- data.frame("KNN_preds" = knn_preds)

svm_pred_df <- data.frame("svm_preds" = svm_preds)

# append the KNN and SVM predictions to the original training dataframe

testing_results <- cbind(testing, knn_preds)

testing_results <- cbind(testing_results, svm_preds)

# select only donor, label, knn_preds, and svm_preds
ml_results <- testing_results%>%select("donor", "label", "knn_preds", "svm_preds")

```


```{r donor_classification}
testing_donors <- unique(ml_results$donor)

donor_classification <- data.frame("donor" = character(),
                                   "num_cells" = numeric(),
                                   "label" = character(), 
                                   "knn_pred" = character(),
                                   "knn_acc" = numeric(),
                                   "knn_correct" = numeric(),
                                   "knn_incorrect" = numeric(),
                                   "svm_pred" = character(),
                                   "svm_acc" = numeric(),
                                   "svm_correct" = numeric(),
                                   "svm_incorrect" = numeric())


# iterate for donors randomly assigned to testing dataset
for (curr_donor in testing_donors) {
  
  # filter for current donor cell labels and predictions
  curr_df <- ml_results%>%filter(donor == curr_donor)
  
  # get number of cells of current donor in testing set
  num_cells <- nrow(curr_df)
  
  # get correct label for current donor
  curr_label = curr_df$label
  
  # summarize current donor classification results for each set of predictions in confusion matrix
  knn_mx = table(curr_df$knn_preds, curr_label)
  print(knn_mx)
  svm_mx = table(curr_df$svm_preds, curr_label)
  print(svm_mx)
  
  knn_correct = NA
  knn_incorrect = NA
  svm_correct = NA
  svm_incorrect = NA
  
  # Extract correct and incorrect classification and calculate accuracy
  if (unique(curr_label) == 1) {
    
    # KNN
    knn_correct = knn_mx[2]
    knn_incorrect = knn_mx[1]
    knn_acc = (knn_mx[2] / (knn_mx[1] + knn_mx[2])) * 100
    
    # SVM
    svm_correct = svm_mx[2]
    svm_incorrect = svm_mx[1]
    svm_acc = (svm_mx[2] / (svm_mx[1] + svm_mx[2])) * 100
  } else if (unique(curr_label) == -1) {
    
    # KNN
    knn_correct = knn_mx[1]
    knn_incorrect = knn_mx[2]
    knn_acc = (knn_mx[1] / (knn_mx[1] + knn_mx[2])) * 100
    
    # SVM
    svm_correct = svm_mx[1]
    svm_incorrect = svm_mx[2]
    svm_acc = (svm_mx[1] / (svm_mx[1] + svm_mx[2])) * 100
  }
  
  
  # predict donor's condition based on cell classification results
  knn_donor_pred = NA
  svm_donor_pred = NA
  
  if (unique(curr_label) == 1 & knn_acc > 50) {
    knn_donor_pred = "Covid" 
  } else {
    knn_donor_pred = "Control"
  }
  
  if (unique(curr_label) == -1 & knn_acc > 50) {
    knn_donor_pred = "Control" 
  } else {
    knn_donor_pred = "Covid"
  }
  
  if (unique(curr_label) == -1 & svm_acc > 50) {
    svm_donor_pred = "Control" 
  } else {
    svm_donor_pred = "Covid"
  }
  
  label = unique(curr_label)
  
  if (label == 1) {
    label = "Covid"
  } else {
    label = "Control"
  }
  
  preds_df <- data.frame(curr_donor, num_cells, label, knn_donor_pred, knn_acc, knn_correct, knn_incorrect, svm_donor_pred, svm_acc, svm_correct, svm_incorrect)
  
  donor_classification <- rbind(donor_classification, preds_df)
}

print(donor_classification)

path <- "./data/"

write.csv(donor_classification, file.path(path, "Donor_Classification_Results.csv"),row.names = TRUE)
```
