---
title: "Extract Count Matrices and Define Training and Testing Data"
author: "Shannon Stiles and Parth Khatri"
date: "Due 12/12/2020"
output: word_document
---

# Notes and Workflow References
Processing Seurat Object: https://satijalab.org/seurat/v3.1/pbmc3k_tutorial.html

```{r setup, include=FALSE}
# setup required libraries
library(devtools)
library(Seurat)
library(cowplot)
library(stringr)

library(tidyseurat)
library(tidyverse)
library(dplyr)
library(patchwork)
library(ggplot2)
library(RColorBrewer)

# PLEASE MAKE SURE YOUR KNIT DIRECTORY IS SET TO YOUR PROJECT DIRECTORY
```

# Read in Merged and Filtered Seurat object
```{r setup_seurat}

# BD Rhapsody scRNA seq
file = "./data/cs760project_COVID19scRNAseq_dataCohort2.rds"

# Read in .rds file to upload Seurat File for Bonn cohort
pbmc <- readRDS(file)

# display seurat object information about Bonn cohort
pbmc

```


# Check memory.size and expand memory.limit to 500 GB, choose n to reach 500 GB
```{r}
curr_memory = memory.size()

```

```{r}

# n = 20 for this curr_memory, will differ by system
memory.limit(size = 20*curr_memory)

```


# Seurat Object Preprocessing

```{r}

# normalize data
pbmc <-NormalizeData(pbmc)
# Find Variable Features for PCA linear dimensionality reduction
pbmc <-FindVariableFeatures(pbmc,selection.method = "vst",nfeatures = 2000)
# scale data
pbmc <-ScaleData(pbmc,features=rownames(pbmc))


```

## Run PCA
```{r}
pbmc <- RunPCA(pbmc, features = VariableFeatures(object = pbmc))
```

# Extract Cell Count Matrices for all Donors within Seurat Object and create Master Dataframe

```{r}
# Example of extracting count matrix for BN-01
# pbmc[,pbmc$donor == "BN-01"]@assays$RNA@counts[,1]

```

# Get all donor IDs
```{r}

# All donor ids were originally used, which included both mild and severe patients, but due to computational efficiency and memory limits, a subset of the 8 severe covid patients and 8 randomly sampled Controls were used instead

# donors <- unique(pbmc@meta.data$donor)
# donors
# 
# sorted_donors <- sort(donors)
# sorted_donors

# 8 severe covid donor and 8 control donor ids used in dataset w

donors = c("BN-11", "BN-12", "BN-14", "BN-15", "BN-16", "BN-17", BN-"18", "BN-19", "BN-22", "BN-23", "BN-24", "BN-26", "BN-27", "BN-28", "BN-29", "BN-31")

```

# Create Dataframes for each donor and bind them together to create master dataframe

```{r}
library(stringr)

# initialize a dataframe to store all counts in
pbmc_df <- data.frame()

for (id in donors) {
  
  # Extract counts matrix for current patient from the RNA assay
  curr_df <- as.data.frame(x = as.matrix(x = GetAssayData(object = pbmc[,pbmc$donor == id])))
  
  # Transpose counts matrix, so features are columns and patient samples are rows
  curr_df <- as.data.frame(t(curr_df))
  
  # find disease label for current sample
  label = unique(pbmc[,pbmc$donor == id]@meta.data$group_per_sample)
  svm_label = 0
  
  
  # assign an svm label for each sample -> Covid Positive = 1, Covid Negative = -1
  if (label == "control") {
    svm_label = -1
  } else if (label == "mild") {
      svm_label = 1 # Unused for final data set of only severe and control donors
  } else if (label == "severe") {
      svm_label = 1
  }
  
  # Add donor id and svm label to current df
  curr_df <- curr_df%>%
    mutate(donor = id,
           label = svm_label)
  
  
  # bind curr_df to pbmc_df
  pbmc_df <- rbind(pbmc_df, curr_df)

  # check to make sure data makes sense
  print(str_c("Patient: ", id, " Label: ", label, " SVM Label: ", svm_label, " Number of Samples: ", nrow(curr_df), " Number of Features: ", ncol(curr_df)))

}


```

# Split by condition to get separate dataframes to randomly sample
```{r}

# severe Covid dataframe
Covid_df <- pbmc_df%>%filter(donor == "BN-11" | donor == "BN-12" | donor == "BN-14" | donor == "BN-15" | donor == "BN-16" | donor == "BN-17" | donor == "BN-18" | donor == "BN-19")

# Control dataframe
Control_df <- first_half_pbmc%>%filter(donor == "BN-22" | donor == "BN-26" | donor == "BN-28" | donor == "BN-31" | donor == "BN-24" | donor == "BN-29" | donor == "BN-23" | donor == "BN-27")


```

# Split Covid and Control subjects randomly into training and testing

## Randomly split donors evenly, seed = 21

```{r}
set.seed(21)

Covid_ids = c("BN-11", "BN-12", "BN-14", "BN-15", "BN-16", "BN-17", "BN-18", "BN-19")

Control_ids = c("BN-22", "BN-26", "BN-28", "BN-31", "BN-24", "BN-29", "BN-23", "BN-27")

Covid_indices = sample(length(Covid_ids), 4)
train_Covid_ids = Covid_ids[Covid_indices]
test_Covid_ids = Covid_ids[-Covid_indices]

Control_indices = sample(length(Control_ids), 4)
train_Control_ids = Control_ids[Control_indices]
test_Control_ids = Control_ids[-Control_indices]

print(str_c("Covid Training Set: ", train_Covid_ids))
print(str_c("Covid Testing Set: ", test_Covid_ids))
print(str_c("Control Training Set: ", train_Control_ids))
print(str_c("Control Testing Set: ", test_Control_ids))


```

## Set up training and testing based on split
```{r}

training = pbmc_df%>%filter(donor == "BN-18" | donor == "BN-11" | donor == "BN-14" | donor == "BN-19" | donor == "BN-26" | donor == "BN-23" | donor == "BN-24" | donor == "BN-28")

testing = pbmc_df%>%filter(donor == "BN-12" | donor == "BN-15" | donor == "BN-16" | donor == "BN-17" | donor == "BN-22" | donor == "BN-31" | donor == "BN-29" | donor == "BN-27")

# save training and testing dataframes to read into other code document
# saveRDS(training, "./data/training_data")
# saveRDS(testing, "./data/testing_data")

```


